---
layout: post
title: " Rendering Rmarkdown  file for a blog post"
output: 
  html_document:
    keep_md: true
author: Nana Boateng
#toc: true
toc_depth: 2
df_print: paged
Time: '`r Sys.time()`'
date: "`r format(Sys.time(), '%B %d, %Y')`"
tags: R ggplot
---



```{r, echo = FALSE}
knitr::opts_chunk$set(
  fig.path = "/Users/nanaakwasiabayieboateng/Documents/memphisclassesbooks/DataMiningscience/rmarkdown",fig.retina=2
)
```


Load Required Packages
---------------------------------

```{r}
pacman::p_load(skimr,kable,kableExtra,tidyverse)
# Load the caret package
library(caret)

# Import dataset
orange <- read.csv('https://raw.githubusercontent.com/selva86/datasets/master/orange_juice_withmissing.csv')

# Structure of the dataframe
str(orange)

# See top 6 rows and 10 columns
head(orange[, 1:10])
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

Creating a Table
---------------------------------


```{r}
library(tidyverse)
library(DT)

kableExtra::kable(orange[1:10, 1:10])

DT::datatable(orange[1:10, 1:10])


```

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file). 

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

```{r}
orange%>%ggplot(aes(PriceMM,color=Purchase))+geom_density()
```