---
title: "Coding Club Markdown Tutorial"
author: "James Bartlett"
date: "17/03/2017"
layout: post
output: html_document
---
#Getting started

First things first, we need to load the packages we need for the rest of the document. 
```{r, message=FALSE}
#Packages
library(dplyr)
```

Now that the package is loaded, we need to load the biodiversity data from the first tutorial.
```{r}
setwd("~/Documents/R Files/Tutorials/R_markdown") #Set the working directory for this example 
edidiv <- read.csv("edidiv.csv")
```

# Calculating the total number of species
### Creating a table of species 

As before, we are going to construct a table of how many species occur in each taxonomic group.
```{r}
richness <- edidiv %>%
    group_by(taxonGroup) %>% #Group by the taxonomy group
    summarise(Species_richness = n_distinct(taxonName)) #Create a count of the number of uniquie names in each taxonomy.
```

We can see that we now have a data frame with two columns. One is for the taxonomic group, and one is for the number of distinct species within each group. We can see a sample of this data here: 
```{r}
head(richness)
```

### Showing the data in a barplot 

Using the table of data, we can now take a look at this in a barplot. 
```{r}
#Create a barplot
barplot(richness$Species_richness, 
        names.arg = richness$taxonGroup, 
        xlab="Taxa", ylab="Number of species", 
        ylim=c(0,600)
        ) 
```

#Most common species 

This time we also want to find the most common species in each taxonomic group.
```{r}
max_abund <- #Calculating the abundance of each species
  edidiv %>%
    group_by(taxonGroup) %>% #Group by taxonomy group 
    summarise(taxonName = names(which.max(table(taxonName)))) #Select the most common species within each taxonomy
```

We now want to join the two data frames together, using "taxonGroup" as the reference.

```{r}
richness_abund <-
inner_join(richness, max_abund, by = "taxonGroup")
richness_abund <- rename(richness_abund, Most_abundant =  taxonName, Taxon = taxonGroup)
```


We can take a final look of what the data frame now looks like.

```{r}
head(richness_abund)
```
